---
title: "Get started with kleaflet.r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with kleaflet.r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(kleaflet.r)
```

A leaflet map with a palette, a legend and popups normally needs a pile of
`addX()` calls, `~column` formulas and `color*()` helpers. `kleaflet()` does
the bookkeeping from a single call, and returns a `kleaflet` object that you
can compose, modify, or turn into a regular leaflet map.

## Input data

`kleaflet()` takes a data frame with longitude and latitude columns (guessed
from names such as `lon`, `lng`, `long`, `longitude`, `lat`, `latitude`; or
given with `lon` and `lat`), an `sf` object (reprojected to WGS84), a
two-column matrix or a named list.

```{r input}
kleaflet(quakes)
kleaflet(data.frame(x = c(2.35, 2.4), y = c(48.85, 48.87)), lon = "x", lat = "y")
```

## Colours, sizes and legends

A column given to `color` is coloured with a palette: a continuous scale for
numbers, a discrete one otherwise. The legend and its title come for free.
`size` and `opacity` rescale a numeric column to a readable range.

```{r colour}
kleaflet(quakes, color = "mag", size = "depth", opacity = "stations",
         size_range = c(2, 12))
```

Numeric colours can be cut in classes with `scale = "bin"` or
`"quantile"`, and `bins`:

```{r bins}
kleaflet(quakes, color = "mag", palette = "YlOrRd", scale = "quantile",
         bins = 4, labels = list(color = "Magnitude"))
```

A palette is a vector of colours (named to fix the colour of each level), a
registered name (`kleaf_palettes()`), the name of an RColorBrewer or viridis
palette, or a function of `n`.

```{r palette}
q <- quakes
q$class <- cut(q$mag, c(4, 5, 6, 7), labels = c("low", "mid", "high"))
kleaflet(q, color = "class",
         palette = c(low = "#2ecc71", mid = "#f1c40f", high = "#e74c3c"))
```

Fixed values are not mappings: wrap colours in `I()`, or give numbers.

```{r fixed}
kleaflet(quakes, color = I("tomato"), size = 3, opacity = 0.4)
```

## Popups and labels

`popup` (on click) and `label` (on hover) accept `TRUE` (all columns), column
names, a template with `{column}`, a function of the data returning HTML, or
one string per row. Values are HTML-escaped.

```{r popups}
kleaflet(q, color = "class",
         popup = "<b>Magnitude {mag}</b><br>Depth: {depth} km",
         label = "{class}")
kleaflet(q, color = "class", popup = c("mag", "depth", "stations"))
```

## Spatial data

With `sf` data the layer type follows the geometry: polygons, lines or
circles. Choropleth maps are one call away.

```{r sf, eval = requireNamespace("sf", quietly = TRUE)}
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
kleaflet(nc, fill = "SID74", palette = "YlOrRd", scale = "bin", bins = 5,
         popup = "<b>{NAME}</b><br>{SID74} cases", theme = "light",
         labels = list(fill = "SID74 (1974)"))
```

## Groups

`group` splits a layer in one overlay per value, with a layers control.

```{r groups}
kleaflet(q, color = "class", group = "class")
```

## Layer types

`kleaf_types()` lists them: `circle` (default for points), `circle_m` (radius
in metres), `marker`, `cluster`, `polygon`, `line`.

```{r types}
kleaf_types()
kleaflet(quakes, type = "cluster", popup = "mag")
```

## Themes: base map and palette

A theme bundles base map tiles and a palette. `tiles` overrides the base map
(several give a switcher), `theme` can also be the name of any provider of
`leaflet::providers`.

```{r themes}
kleaf_themes()
kleaflet(q, color = "class", theme = "inrae", title = "Fiji earthquakes",
         caption = "Source: datasets::quakes")
kleaflet(q, tiles = c("CartoDB.Positron", "Esri.WorldImagery"))
```

Add your own with `kleaf_register_theme()` and `kleaf_register_palette()`;
`options(kleaflet.theme = "inrae")` sets a session default.

## Composing maps

`+` adds the layers of another `kleaflet`; colours and legends are shared.
Settings of the right-hand side (theme, titles, view...) win.

```{r compose}
a <- data.frame(lon = c(2.35, 2.40), lat = c(48.85, 48.87))
b <- data.frame(lon = c(2.30, 2.33), lat = c(48.84, 48.88))
kleaflet(a, color = "Stations") + kleaflet(b, color = "Sensors")
```

Pipe-friendly modifiers return a modified `kleaflet`: `kleaf_add()`,
`kleaf_labs()`, `kleaf_theme()`, `kleaf_tiles()`, `kleaf_palette()`,
`kleaf_legend()`, `kleaf_view()` and `kleaf_controls()`.

```{r modify}
kleaflet(q, color = "mag") |>
  kleaf_add(type = "marker", popup = "mag") |>
  kleaf_labs(title = "Fiji", color = "Magnitude") |>
  kleaf_palette("Reds", scale = "bin", bins = 4) |>
  kleaf_legend("bottomleft") |>
  kleaf_view(c(176, -20, 5)) |>
  kleaf_controls(scale_bar = TRUE, minimap = TRUE)
```

A function of the map, added with `+`, runs last: it is the escape hatch for
anything leaflet (or one of its extensions) can do.

```{r extras}
kleaflet(quakes, color = "mag") + (function(m) leaflet::addScaleBar(m))
```

## Going back to leaflet

`as_leaflet()` returns the regular leaflet map, to continue with `|>`, to use
in Shiny (`leaflet::renderLeaflet(as_leaflet(p))`) or to save, with
`kleaf_save(p, "map.html")`.

```{r as_leaflet}
as_leaflet(kleaflet(quakes, color = "mag")) |>
  leaflet::addCircleMarkers(lng = 178, lat = -20, radius = 20, color = "red")
```

## Extending kleaflet.r

* **Input classes**: define `as_kdata.<class>()` returning a data frame or an
  `sf` object. It may set the attributes `kleaf_lon`, `kleaf_lat` and
  `kleaf_type` as defaults.
* **Layer types**: `kleaf_register_type(name, fun, geometry, aes, params)`
  says which leaflet function to call and which of its arguments each
  aesthetic sets.
* **Themes and palettes**: `kleaf_register_theme()` and
  `kleaf_register_palette()`.

```{r extend}
kleaf_register_type(
  "big_dot", "addCircleMarkers", aes = c(colour = "color"),
  params = list(radius = 12, fillOpacity = 0.9)
)
kleaflet(quakes, color = "mag", type = "big_dot")
```
